---
title: "illumina data"
output: html_document
---


# loading libraries

```{r setup, include=FALSE}
library(hdf5r)
library(SingleCellExperiment)
library(biomaRt)
library(Seurat)
library(dplyr)
library(singleCellTK)
library(Seurat)
library(scater)
library(singleCellTK)
library(SCHNAPPs)
library(iSEE)
```

# getFeatureDataSummary

internal function used to get annotations from ENSEMBL

```{r getFeatureDataSummary}
getFeatureDataSummary <- function(ver = 0, dataset = "mmusculus_gene_ensembl", gbmMat = gbmMat) {
  if (ver > 0) {
    ensemblOld <- useEnsembl(biomart = "ensembl", version = ver, dataset = dataset)
  } else {
    ensemblOld <- useEnsembl(biomart = "ensembl", dataset = dataset)
  }

  if (!startsWith(rownames(gbmMat)[1], "ENS")) {
    vals <- rownames(gbmMat)
    filters <- "external_gene_name"
  } else {
    vals <- rownames(gbmMat)
    filters <- "ensembl_gene_id"
  }
  featureData <- tryCatch(
    getBM(
      attributes = c(
        "ensembl_gene_id",
        "external_gene_name",
        "description",
        "chromosome_name",
        "genomic_coding_start",
        "genomic_coding_end",
        "gene_biotype"
      ),
      filters = filters,
      values = vals,
      mart = ensemblOld
    ),
    error = function(cond) {
      return("biomart connection error")
    },
    warning = function(cond) {
      return("Biomart warning")
    }
  )
  featureData$genomic_coding_start[is.na(featureData$genomic_coding_start)] <- 0
  featureData$genomic_coding_end[is.na(featureData$genomic_coding_end)] <- 0
  featureData_summary <- featureData %>%
    group_by(ensembl_gene_id, external_gene_name, description, chromosome_name, gene_biotype) %>%
    summarize(
      min_genomic_coding_start = min(genomic_coding_start, na.rm = TRUE),
      max_genomic_coding_end = max(genomic_coding_end, na.rm = TRUE)
    )
  featureData_summary <- as.data.frame(featureData_summary)
  rownames(featureData_summary) <- featureData_summary$ensembl_gene_id
  featureData_summary <- featureData_summary[, -1]

  colnames(featureData_summary) <- c(
    "Associated.Gene.Name",
    "Description",
    "Chromosome.Name",
    "Gene.Biotype",
    "Gene.Start..bp.",
    "Gene.End..bp."
  )
  featureData_summary <- featureData_summary[, c(
    "Description",
    "Chromosome.Name",
    "Gene.Start..bp.",
    "Gene.End..bp.",
    "Associated.Gene.Name",
    "Gene.Biotype"
  )]
  return(featureData_summary)
}
```



# load data

download from

https://support.10xgenomics.com/single-cell-gene-expression/datasets

(has already been done)

* 5k Peripheral blood mononuclear cells (PBMCs) from a healthy donor with cell surface proteins (v3 chemistry)

** Feature / cell matrix HDF5 (filtered)	16.34 MB	3366a47283177fe9af143d5819fad61f

## load from h5 file
```{r h5 data}
if (Sys.info()["sysname"] == "Darwin") {
  h5file <- "~/Downloads/5k_pbmc_protein_v3_filtered_feature_bc_matrix.h5"
} else {
  h5file <- "~/../Downloads/5k_pbmc_protein_v3_filtered_feature_bc_matrix.h5"
}
scexSeurat <- Seurat::Read10X_h5(h5file, use.names = TRUE, unique.features = TRUE)
names(scexSeurat)
save(file = "scexSeurat.RData", list = c("scexSeurat"))
```


## load from directory 

here you would potentially put other ways of importing data
```{r empty}

```


## rename samples

## combine samples

# process

```{r schnapps}
# pd holds the meta data for the cells
pd <- data.frame(
  barcodes = sub("(.*)-(.*)", "\\1", colnames(scexSeurat[[1]])),
  sampleNames = sub(".*-(.*)", "\\1", colnames(scexSeurat[[1]]))
)
# barcodes is a name that is needed by SCHNAPPs
pd$barcodes <- as.character(pd$barcodes)
rownames(pd) <- colnames(scexSeurat[[1]])

# scexSeurat[[2]]: the cytometry data
pd <- cbind(pd, as.data.frame(t(scexSeurat[[2]])))

# we also create and load the log transformed version for visualization purposes.
logdf <- log(as.data.frame(t(scexSeurat[[2]])))
logdf <- do.call(data.frame, lapply(logdf, function(x) replace(x, is.infinite(x), 0)))
colnames(logdf) <- paste0("lg_", colnames(logdf))
pd <- cbind(pd, logdf)
str(pd)

# remove genes with zero counts over all cells
scexSeurat[[1]] <- scexSeurat[[1]][-which(rowSums(scexSeurat[[1]]) == 0), ]

# load some additional meta data for genes from ENSEMBL
if (!file.exists("feature93.RData")) {
  featureData_summary93 <- getFeatureDataSummary(ver = 93, dataset = "hsapiens_gene_ensembl", gbmMat = scexSeurat[[1]])
  save(file = "feature93.RData", list = ("featureData_summary93"))
} else {
  load("feature93.RData")
}
featureData_summary <- featureData_summary93
# genes not in annotation.
rownames(scexSeurat[[1]])[!rownames(scexSeurat[[1]]) %in% featureData_summary$Associated.Gene.Name ]

featuredata <- featureData_summary[featureData_summary$Associated.Gene.Name %in% rownames(scexSeurat[[1]]), ]

# remove duplicates
featuredata <- featuredata[!duplicated(featuredata$Associated.Gene.Name), ]

# only take genes where we have descriptions
# matrix requires numeric index
scIdx <- which(rownames(scexSeurat[[1]]) %in% featuredata$Associated.Gene.Name)
scexSeurat[[1]] <- scexSeurat[[1]][scIdx, ]

rownames(featuredata) <- featuredata$Associated.Gene.Name

# check that  rownames are included in both
sum(!rownames(featuredata) %in% rownames(scexSeurat[[1]]))
sum(!rownames(scexSeurat[[1]]) %in% rownames(featuredata))

# reorder rows
featuredata <- featuredata[rownames(scexSeurat[[1]]), ]
featuredata$id <- featuredata$Associated.Gene.Name
featuredata$symbol <- featuredata$Associated.Gene.Name
colnames(pd)[colnames(pd) == "barcodes"] <- "barcode"
# scEx - rowData doesn't contain id and/or symbol columns
# ×
# featuredata - one of is missing: symbol, Description)
scEx <- SingleCellExperiment(
  assay = list(counts = scexSeurat[[1]]),
  colData = pd,
  rowData = featuredata
)

outfile <- paste0(basename(h5file), ".RData")

save(file = outfile, list = c("scEx"))

# just a subset randomly sampled
scExSmall <- scEx[, sample(1:ncol(scEx), 1111)]

outfile <- paste0(basename(h5file), "small.RData")

save(file = outfile, list = c("scExSmall"))


scEx <- normalizeSCE(scEx,
  exprs_values = "counts", return_log = TRUE,
  log_exprs_offset = NULL, centre_size_factors = TRUE,
  preserve_zeroes = FALSE
)

scExS <- as.Seurat(scEx,
  project = "SingleCellExperiment"
)

save.image(file = "perpIllumina.RData")
```


```{r calculateQCMetrics}
# load("perpIllumina.RData")
ave_counts <- calculateAverage(scEx,
  exprs_values = "counts",
  use_size_factors = TRUE, subset_row = NULL,
  BPPARAM = SerialParam()
)

cpm(scEx) <- calculateCPM(scEx,
  exprs_values = "counts", use_size_factors = TRUE,
  subset_row = NULL
)

scEx <- calculateQCMetrics(scEx)
```

```{r seurat}
pbmc <- CreateSeuratObject(counts = assays(scEx)[["counts"]], project = "scCourse2019", min.cells = 3, min.features = 200)

save(file = "pbmc5k.seurat.RData", "pbmc")
```

## singleCellTK

https://compbiomed.github.io/sctk_docs/articles/v01-Introduction_to_singleCellTK.html

```{r singleCellTK}

# singleCellTK(scEx)
```


## iSEE

https://bioconductor.org/packages/release/bioc/html/iSEE.html

```{r iSEE}
# iSEE(scEx)

```


## SCHNAPPs

https://github.com/C3BI-pasteur-fr/UTechSCB-SCHNAPPs
https://c3bi-pasteur-fr.github.io/UTechSCB-SCHNAPPs/


```{r}
# schnapps()

```

